---
title: "Unemployment Data Analysis"
author: "Ryan Honea"
date: "4/24/2018"
header-includes:
   - \institute{Austin Peay State University}
   - \logo{\includegraphics[scale=.05]{AP}}
output: 
  beamer_presentation:
    theme: "CambridgeUS"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library("knitr")
library("dplyr")
library("readr")
library("quantmod")
library("TSA")
library("astsa")
library("fGarch")
library("forecast")
library("car")
x <- read_csv("data7.csv")
x <- ts(x[,2])
```

## Initial Plotting
```{r, echo = FALSE}
plot(x, ylab = "Unemployment Rate", xlab = "Time Index",
     main = "Unemployment Data 01/01/1948 - 03/01/2018")
```

## Time Series Display
```{r, echo = FALSE}
tsdisplay(x)
```

## Analysis on Plots

* The Unemployment Data covers January 1948 to March 2018 and thus covers many times of uncertainty
* Jumps from postwar uncertainty, 1980s recessions, 2000s recession
* These periods of volatility lend itself to using an ARMA GARCH model.

## Possible Models
* Possible model are
    * ARMA(1,1) $\times$ Garch (1,1)
    * ARMA(1,1) $\times$ Garch (1,2)
    * ARMA(1,1) $\times$ Garch (2,1)
    * ARMA(1,1) $\times$ Garch (2,2)
* Conditional Distribution used will be skewed student t-distriution

## Model Input
```{r, eval = FALSE, echo = TRUE}
fit1 <- garchFit(~arma(1,1) + garch(1,1), 
                 x, cond.dist = "sstd")
fit2 <- garchFit(~arma(1,1) + garch(1,2), 
                 x, cond.dist = "sstd")
fit3 <- garchFit(~arma(1,1) + garch(2,1), 
                 x, cond.dist = "sstd")
fit4 <- garchFit(~arma(1,1) + garch(2,2), 
                 x, cond.dist = "sstd") 
```

```{r, echo = F, results = "hide", warning = F}
fit1 <- garchFit(~arma(1,1) + garch(1,1), x, cond.dist = "sstd")
fit2 <- garchFit(~arma(1,1) + garch(1,2), x, cond.dist = "sstd")
fit3 <- garchFit(~arma(1,1) + garch(2,1), x, cond.dist = "sstd")
fit4 <- garchFit(~arma(1,1) + garch(2,2), x, cond.dist = "sstd") 
```

## Residual Analysis Fit1 QQ SSTD Plot
```{r, echo = FALSE}
plot(fit1, which = 13, main = "QQ SSTD Plot for Fit1")
```


## Residual Analysis Fit1 ACF
```{r, echo = FALSE}
plot(fit1, which = 11, main = "ACF of Squared Standardized Residuals Fit1")
```

## Residual Analysis Fit2 QQ SSTD Plot
```{r, echo = FALSE}
plot(fit2, which = 13, main = "QQ SSTD Plot for Fit2")
```


## Residual Analysis Fit2 ACF
```{r, echo = FALSE}
plot(fit2, which = 11, main = "ACF of Squared Standardized Residuals Fit2")
```

## Residual Analysis Fit3 QQ SSTD Plot
```{r, echo = FALSE}
plot(fit3, which = 13, main = "QQ SSTD Plot for Fit3")
```


## Residual Analysis Fit3 ACF
```{r, echo = FALSE}
plot(fit3, which = 11, main = "ACF of Squared Standardized Residuals Fit3")
```

## Residual Analysis Fit4 QQ SSTD Plot
```{r, echo = FALSE}
plot(fit4, which = 13, main = "QQ SSTD Plot for Fit3")
```


## Residual Analysis Fit4 ACF
```{r, echo = FALSE}
plot(fit4, which = 11, main = "ACF of Squared Standardized Residuals Fit3")
```

## Final Decision
Based on Ljung-Box values on Squared Residuals, I chose fit4 which had the best Ljung-Box Test values, the best LM Arch test values, and the best looking residuals.

Coefficients are on following slide

## Coefficients of ARMA(1,1) GARCH(2,2)

```{r, echo = FALSE, results = "markup"}
knitr::kable(fit4@fit$matcoef)
```


## Forecasted Data
```{r, echo = FALSE}
fcast1=predict(fit4,n.ahead=30,plot=TRUE)
```