---
title: "Using Simulation to Present Complex Topics"
author: "Ryan Honea"
institution: "Austin Peay State University"
date: "9/30/2017"
output: 

  beamer_presentation:
    template: conftemplate.tex
    slide_level: 3
    theme: "Antibes"
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```


# Introduction: Why Simulate?
## Why Simulate?
### Why Simulate?
Common Questions:

- Is it worthwhile to teach programming?
- Wouldn't writing a simulation be hard?
- Why not just use theorems and proofs and other such nonsense?

## Motivation
### Why We Love Math

The Big Questions
- Why do we love math?
- What drives students to love math?
- How can we utilize this information to enhance student's learning?

### Why We Love Math: GRAPHS
```{r, echo = FALSE}
z = function(n,c,p) {
  val = (c - n*p)/(sqrt(n*p*(1-p)))
  return(val)
}
probabilities = c(.600, .650, .700)
n = seq(1,200, by = 1)
power = matrix(NA, nrow = 200, ncol = 1)
z_val = matrix(NA, nrow = 200, ncol = 1)
prob = matrix(NA, nrow = 200, ncol = 1)
c = matrix(NA, nrow = 200, ncol = 1)
dfs <- list()
for (j in 1:3) {
  for (i in 1:200) {
    c[i,] = qnorm(.975, n[i]*.5, sqrt(n[i]*.5*(1-.5)))
    z_val[i,] = z(n[i], c[i,], probabilities[j])
    power[i,] = (1 - pnorm(z_val[i,], 0, 1))
    prob[i,] = probabilities[j]
  }
  dfs[[j]] <- data.frame(n, power, prob)
}
plot(dfs[[1]]$n, dfs[[3]]$power,
     main = "Power versus Number of Samples for Sign Test",
     xlab = "Number of Samples", ylab = "Power",
     col = "darkgreen", type = "l", lty=3)
lines(dfs[[1]]$n, dfs[[2]]$power, col = "dodgerblue", lty = 2)
lines(dfs[[1]]$n, dfs[[1]]$power, col = "firebrick4", lty = 1)
legend(1, legend=c("p = .60", "p = .65", "p = .70"),
       col=c("firebrick4","dodgerblue","darkgreen"), lty=1:3, cex=0.8)
```
### Why We Love Math: FUNctions
### Why We Love Math: It Just Works?

### Senior Research Project

## Context of this Presentation
### The "Traditional" Statistics Course

### More Intense Problems

# The "Traditional" Statistics Course
## Measures of Center
## Sampling
## Regression
## Hypothesis Testing

# Elevating the Statistics Course
## Non-Normal Distributions
## Central Limit Theorem

# Conclusion
## Conclusion
### Conclusion

## Acknowledgement
### Acknowledgements