---
title: "Chapter Nine"
author: "Ryan Honea"
date: "10/9/2017"
header-includes:
   - \usepackage{bbm}
   - \usepackage{bm}
   - \newcommand{\infmat}{\bm{\mathcal{J}}}
   - \usepackage{booktabs}
   - \usepackage{multirow}
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
require(ggplot2)
```

## Exercise Two
#### Question
The data in the table below (Table 9.13) are numbers of insurance policies, $n$, and numbers of claims, $y$, for cars in various insurance categories, $CAR$, tabublated by age of policy holder, $AGE$, and district where the policy holder lived ($DIST = 1$, for London and other major cities, and $DIST = 0$, otherwise). The table is derived from the $CLAIMS$ data set in Aitkin et al. (2005) obtained from a paper by Baxter et al. (1980).

\begin{center}
\begin{tabular}{@{}ccrrlrr@{}}
\toprule
      &       & \multicolumn{2}{c}{DIST = 0}                      &  & \multicolumn{2}{c}{DIST = 1}                      \\ \cmidrule(lr){3-4} \cmidrule(l){6-7} 
$CAR$ & $AGE$ & \multicolumn{1}{c}{$y$} & \multicolumn{1}{c}{$n$} &  & \multicolumn{1}{c}{$y$} & \multicolumn{1}{c}{$n$} \\ \midrule
1     & 1     & 65                      & 317                     &  & 2                       & 20                      \\
1     & 2     & 65                      & 476                     &  & 5                       & 33                      \\
1     & 3     & 52                      & 486                     &  & 4                       & 40                      \\
1     & 4     & 310                     & 3259                    &  & 36                      & 316                     \\
2     & 1     & 98                      & 486                     &  & 7                       & 31                      \\
2     & 2     & 159                     & 1004                    &  & 10                      & 81                      \\
2     & 3     & 175                     & 1355                    &  & 22                      & 122                     \\
2     & 4     & 877                     & 7660                    &  & 102                     & 724                     \\
3     & 1     & 41                      & 223                     &  & 5                       & 18                      \\
3     & 2     & 117                     & 539                     &  & 7                       & 39                      \\
3     & 3     & 137                     & 697                     &  & 16                      & 68                      \\
3     & 4     & 477                     & 3442                    &  & 63                      & 344                     \\
4     & 1     & 11                      & 40                      &  & 0                       & 3                       \\
4     & 2     & 35                      & 148                     &  & 6                       & 16                      \\
4     & 3     & 39                      & 214                     &  & 8                       & 25                      \\
4     & 4     & 167                     & 1019                    &  & 33                      & 114                     \\ \bottomrule
\end{tabular}
\end{center}

#### Solution
Below is just data entry.
```{r, echo = TRUE}
df <- data.frame(
  CAR = CAR <- rep(rep(1:4, each = 4),2),
  AGE = AGE <- rep(rep(1:4, 4),2),
  DIST = DIST <- rep(0:1, each = 16),
  y = y <- c(  65,  65,  52, 310,  98, 159, 175, 877,
               41, 117, 137, 477,  11,  35,  39, 167,
                2,   5,   4,  36,   7,  10,  22, 102,
                5,   7,   16, 63,   0,   6,   8,  33),
  n = n <- c( 317, 476, 486,3259, 486,1004,1355,7660,
              223, 539, 697,3442,  40, 148, 214,1019,
               20,  33,  40, 316,  31,  81, 122, 724,
               18,  39,  68, 344,   3,  16,  25, 114)
)
```

**(a)**: Calculate the rate of claims $y/n$ for each category and plot the rates by $AGE$, $CAR$, and $DIST$ to get an idea of the main effects of those factors.

*Solution: *
```{r, echo = TRUE}
df$rate <- df$y/df$n
par(mfrow = c(1,3))
plot(df$DIST, df$rate, xlab= "District", ylab = "Claims per Insurance Policy")
plot(df$CAR, df$rate, xlab= "Insurance Category", ylab = "Claims per Insurance Policy")
plot(df$AGE, df$rate, xlab= "Age of Policy Holder", ylab = "Claims per Insurance Policy")
```

Based on these plots, it appears that the difference between districts is just an increased variance of rates in lower population areas. Insurance categories from 1 to 4 seem to have increasing rates as well as increasing variance per rate level. From the youngest age group to the second youngest, there is an increase, but a decrease following that in claim rates.

**(b)**: Use Poisson regression to estimate the main effects (each treated as categorical and modeled using indicator variables) and interaction terms.

*Solution: * First, we convert ```CAR```, ```AGE```, and ```DIST``` into factors so that they are treated categorically.

```{r, echo = TRUE}
df$CAR <- factor(df$CAR); df$AGE <- factor(df$AGE); df$DIST <- factor(df$DIST)
poismod <- glm(y ~ CAR*AGE*DIST + offset(log(n)), data = df,
               family = poisson(link = "log"))
summary(poismod)
```

**(c)**: Based on the modelling in (b), Aitkin et al. (2005) determined that all the interactions were unimportant and decided that $AGE$ and $CAR$ could could be treated as though they were continuous variables. Fit a model incorpororating these features and compare it with the best model in (b). What conclusions do you reach?

*Solution: * First, we convert ```CAR``` and ```AGE``` into numeric values and then create the other model.

```{r, echo = TRUE}
df$CAR <- as.numeric(df$CAR); df$AGE <- as.numeric(df$AGE)
poismod2 <- glm(y ~ CAR + AGE + DIST + offset(log(n)), data = df,
                family = poisson(link = "log"))
summary(poismod2)
```

We can check whether or not this model fits well by comparing it to the saturated model which was found in (b). This is the residual deviance noted in the summary.

```{r, echo = TRUE}
p <- pchisq(poismod2$deviance, 28)
cat("p-value: ", p)
```

Based on the p-value, there is not enough evidence to suggest that this model does not fit the data and so we would prefer to use this simpler model.

\pagebreak

## Exercise Three
#### Question
This question relates to the flu vaccine trial data in the table below (Table 9.6).

\begin{center}
\begin{tabular}{@{}lcccc@{}}
\toprule
         & \multicolumn{4}{c}{Response}     \\
         & Small & Moderate & Large & Total \\ \midrule
Placebo & 25    & 8        & 5     & 38    \\
Vaccine  & 6     & 18       & 11    & 35    \\ \bottomrule
\end{tabular}
\end{center}

#### Solution
Below is data entry.
```{r, echo = TRUE}
vacctrial <- as.table( matrix(
  c(25, 8, 5, 6, 18, 11), byrow = TRUE, ncol = 3)
)
colnames(vacctrial) <- c("Small", "Moderate", "Large")
rownames(vacctrial) <- c("Vaccine", "Placebo")
```

**(a)**: Using a conventional chi-squared test and an appropriate log-linear model, test the hypothesis that the distribution of responses is the same for the placebo and vaccine groups.

*Solution: * The chi-squared test is performed below.
```{r, echo = TRUE}
require(MASS)
chisq.test(vacctrial)
```

With at least 95% confidence, we reject the null hypothessis that the distribution of the responses are the same for the placebo and vaccine groups. We test this again using the additive log-linear model.

```{r, echo = TRUE}
df <- as.data.frame(vacctrial)
colnames(df) <- c("Treatment", "Response", "Freq")
df$Treatment <- relevel(df$Treatment, ref = "Placebo")
df$Response <- relevel(df$Response, ref = "Small")
poismod <- glm(Freq ~ Treatment + Response, data = df,
               family = poisson(link = "log"))
fits <- c(fitted.values(poismod))
X2 <- sum(((df$Freq - fits)/sqrt(fits))^2)
1 - pchisq(X2, 2)
```

With an equal p-value, we make the same conclusion.

**(b)**: For the model corresponding to the hypothesis of homogeneity of response distributions, calculate the fitted values, the Pearson and deviance residuals, and the goodness of fit statistics $\chi^2$ and $D$. Which of the cells of the table contribute most to $\chi^2$ (or $D$)? Explain and interpret these results.

*Solution: *  We test the hypothesis of homogeneity by comparing the saturated and the additive model.

```{r, echo = TRUE}
poissat <- glm(Freq ~ Treatment*Response, data = df,
               family = poisson(link = "log"))
dev.res <- sign(df$Freq - fits) *(sqrt(2*(df$Freq*log(df$Freq/fits)-(df$Freq-fits))))
deviance <- sum(dev.res^2)
deviance
2*(logLik(poissat) - logLik(poismod))
```

These give the same results as before and so we see homogeneity of response distribution.

**(c)**: Re-analyze these data using ordinal logistic regression to estimate cut-points for for a latent continuous response variable and to estimate a location shift between the two treatment groups. Sketch a rough diagram to illustrate the model which forms the conceptual base for this analysis (see Exercise 8.4).

*Solution: * 
```{r, echo = TRUE}
require(nnet)
ordmod <- polr(Response ~ Treatment, weights = Freq, data = df)
df$probs <- predict(ordmod, type = "probs")
df[1:2,c("Treatment", "probs")]
shift <- log((df[2,4][1]*(df[1,4][2]+df[1,4][3]))/(df[1,4][1]*(df[2,4][2]+df[2,4][3])))
cat("Estimated Shift:", shift)
```