---
title: "Chapter Seven"
author: "Ryan Honea"
date: "10/9/2017"
header-includes:
   - \usepackage{bbm}
   - \usepackage{bm}
   - \newcommand{\infmat}{\bm{\mathcal{J}}}
   - \usepackage{booktabs}
   - \usepackage{multirow}
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
require(ggplot2)
```

## Exercise One
#### Question
The numebr of deaths from leukemia and other cancers among survivors of the Hiroshima atom bomb are shown in Table 7.12 (below), classified by the radiation dose received. The data refers to deaths during the period 1950-1959 among survivors who were aged 25 to 64 years in 1950 (from data set 13 of Cox and Snell 1981, attributed to Otake 1979).

\begin{center}
\begin{tabular}{@{}lcccccc@{}}
\toprule
              & \multicolumn{6}{c}{Radiation Dose (rads)}  \\
Deaths        & 0   & 1-9 & 10-49 & 50-99 & 100-199 & 200+ \\ \midrule
Leukemia      & 13  & 5   & 5     & 3     & 4       & 18   \\
Other Cancers & 378 & 200 & 151   & 47    & 31      & 33   \\ \midrule
Total Cancers & 391 & 205 & 156   & 50    & 35      & 51   \\ \bottomrule
\end{tabular}
\end{center}

#### Solution
**(a)**: Obtain a suitable model to describe the dose-response relationship between radiation and the proportional cancer mortality rates for leukemia.

*Solution: * First we enter the data and plot it to gain an initial understanding of our data.
```{r, echo = TRUE}
Dosage <- c("0", "1-9", "10-49","50-99","100-199","200+")
Dose.Bounded <- c(0,1,10,50,100,200)
Leukemia <- c(13, 5, 5, 3, 4, 18)
Other <- c(378, 200, 151, 47, 31, 33)
labels <- c("Leukemia","Other")
colors <- c(rep("red",6), rep("blue",6))
plot(c(Dose.Bounded, Dose.Bounded), c(Leukemia, Other), 
     pch = c(16,16), col = colors,
     xlab = "Dosage Lower Bound", ylab = "Deaths",
     main = "Cancer Deaths by Dosage in Rads")
legend("topright", title = "Cancer Type",
       labels, col = c("red","blue"), pch = c(16,16))
```

There is clearly a divide here that we can hope to find with a logistic model.

```{r, echo = TRUE}
cancers <- cbind(Leukemia, Other)
logmod <- glm(cancers ~ Dose.Bounded, family=binomial(link = "logit"))
logmod
```

**(b)**: Examine how well the model describes the data.

*Solution*: Below is a graph checking the accuracy of the model by drawing a line along the proportion of cancer deaths.

```{r, echo = FALSE}
proportions <- Leukemia/Other
plot(Dose.Bounded, proportions, 
     pch = c(16,16), xlab = "Dosage Lower Bound", 
     ylab = "Proportion of Cancer Deaths (# Leukemia / # Other)",
     main = "Cancer Deaths by Dosage in Rads")
b0 <- logmod$coefficients[[1]]
b1 <- logmod$coefficients[[2]]
y <- function(x) {
  exp(b0 + b1*x) /
  (1 + exp(b0 + b1*x))
} 
curve(y, add=TRUE)
```

As there are six covariate patterns and one parameter, we can find the Deviance and check it against $\chi^2(6-1)$.
```{r, echo = TRUE}
deviance = logmod$deviance
p.value <- 1 - pchisq(deviance, df = 5)
cat("P.Value:",p.value)
```

Conversely, one can use the Hosmer Lemeshow test where $D \sim \chi^2(g-2)$ where in this case, we have six groups, so $g = 6$.

```{r, echo = TRUE}
deviance = logmod$deviance
p.value <- 1 - pchisq(deviance, df = 4)
cat("P.Value:",p.value)
```
**(c)**: Interpret the results.

*Solution*: Based on the above p-values, it is strongly suggested that this is a strong model with a good fit.


\pagebreak

## Exercise Two
#### Question
**Odds ratios**. Consider a $2\times2$ contingency table from a prospective study in which people who were or were not exposed to some pollutants are followed up and, after several years, categorized according to the presence or absence of a disease. Table 7.13 (below) shows the probabilities for each cell. The odds of disease for either exposure group is $O_i = \pi_i/(1-\pi_i)$, for $i = 1, 2,$ and so the odds ratio
$$
\phi = \frac{O_1}{O_2} = \frac{\pi_1(1-\pi_2)}{\pi_2(1-\pi_1)}
$$

is a measure of the relative likelihood of disease for the exposed and not exposed groups.

\begin{center}
\begin{tabular}{@{}lcc@{}}
\toprule
            & Diseased & Not diseased \\ \midrule
Exposed     & $\pi_1$  & $1-\pi_1$    \\
Not exposed & $\pi_2$  & $1-\pi_2$    \\ \bottomrule
\end{tabular}
\end{center}

#### Solution
**(a)**: For the simple logistic model $\pi_i = e^{\beta_i}/(1+e^{\beta_i})$, show that if there is no difference between the exposed and not exposed groups (i.e. $\beta_1 = \beta_2$), then $\phi = 1$.

*Solution*: If $\beta_1 = \beta_2$, then
$$\pi_1 = \frac{e^{\beta_1}}{1 + e^{\beta_1}} = \frac{e^{\beta_2}}{1 + e^{\beta_2}} = \pi_2$$
and thus
$$\phi = \frac{\pi_1(1 - \pi_2)}{\pi_2(1-\pi_1)} = \frac{\pi_1(1-\pi_1)}{\pi_1(1-\pi_1)} = 1$$

**(b)**: Consider $J$ $2\times2$ tables like Table 7.13, one for each level $x_j$ of a factor, such as age group, with $j = 1,...,J$. For the logistic model
$$
\pi_{ij} = \frac{\exp(\alpha_i + \beta_ix_j)}{1 + \exp(\alpha_i + \beta_ix_j)}, \quad i = 1,2, \quad j = 1,...,J.
$$

Show that $\log\phi$ is constant over all table if $\beta_1 = \beta_2$ (McKinlay 1979).

*Solution:* If $\beta_1 = \beta_2$ and $\eta_{i} = \alpha_i + \beta_ix_j$
\begin{align*}
\phi_j  &= \frac{\pi_{1j}(1-\pi_{2j})}{\pi_{2j}(1-\pi_{1j})}
         = \dfrac{\frac{\exp(\eta_1)}{1 + \exp(\eta_1)}\left(1 - \frac{\exp(\eta_2)}{1 + \exp(\eta_2)}\right)}
            {\frac{\exp(\eta_2)}{1 + \exp(\eta_2)}\left(1 - \frac{\exp(\eta_1)}{1 + \exp(\eta_1)}\right)}\\
        &= \dfrac{\frac{\exp(\eta_1)}{1 + \exp(\eta_1)}\left(\frac{1}{1 + \exp(\eta_2)}\right)}
            {\frac{\exp(\eta_2)}{1 + \exp(\eta_2)}\left(\frac{1}{1 + \exp(\eta_1)}\right)}
         = \frac{\eta_1}{\eta_2}\\
        &= \exp(\eta_1 - \eta_2) = \exp(\alpha_1 - \alpha_2 - x_j(\beta_1 - \beta_2))\\
        &= \exp(\alpha_1 - \alpha_2 - x_j(\beta_1 - \beta_1)) = \exp(\alpha_1 - \alpha_2)
\end{align*}

So we have shown that $\phi_j$ is constant when $\beta_1 = \beta_2$.

\pagebreak

## Exercise Three
#### Question
Tables 7.14 and 7.15 (both below) show the survival 50 years after graduation of men and women who graduated each year from 1938 and 1947 from various Faculties of the University of Adelaide (data compiled by J.A. Keats). The columns labeled $S$ contain the number of graduates who survived and the columns labeled $T$ contain the total number of graduates. There were insufficient women graduates from the Faculties of Medicine and Engineering to warrant analysis.

\begin{center}
\begin{tabular}{@{}crrrrrrrr@{}}
\toprule
Year       & \multicolumn{8}{c}{Faculty}                                                                                                                                                                   \\
of         & \multicolumn{2}{c}{Medicine}                  & \multicolumn{2}{c}{Arts}                      & \multicolumn{2}{c}{Science}                   & \multicolumn{2}{c}{Engineering}               \\
Graduation & \multicolumn{1}{c}{S} & \multicolumn{1}{c}{T} & \multicolumn{1}{c}{S} & \multicolumn{1}{c}{T} & \multicolumn{1}{c}{S} & \multicolumn{1}{c}{T} & \multicolumn{1}{c}{S} & \multicolumn{1}{c}{T} \\ \midrule
1938       & 18                    & 22                    & 16                    & 30                    & 9                     & 14                    & 10                    & 16                    \\
1939       & 16                    & 23                    & 13                    & 22                    & 9                     & 12                    & 7                     & 11                    \\
1940       & 7                     & 17                    & 11                    & 25                    & 12                    & 19                    & 12                    & 15                    \\
1941       & 12                    & 25                    & 12                    & 14                    & 12                    & 15                    & 8                     & 9                     \\
1942       & 24                    & 50                    & 8                     & 12                    & 20                    & 28                    & 5                     & 7                     \\
1943       & 16                    & 21                    & 11                    & 20                    & 16                    & 21                    & 1                     & 2                     \\
1944       & 22                    & 32                    & 4                     & 10                    & 25                    & 31                    & 16                    & 22                    \\
1945       & 12                    & 14                    & 4                     & 12                    & 32                    & 38                    & 19                    & 25                    \\
1946       & 22                    & 34                    &                       &                       & 4                     & 5                     &                       &                       \\
1947       & 28                    & 37                    & 13                    & 23                    & 25                    & 31                    & 25                    & 35                    \\
Total      & 177                   & 275                   & 92                    & 168                   & 164                   & 214                   & 103                   & 142                   \\ \bottomrule
\end{tabular}

\begin{tabular}{@{}crrrr@{}}
\toprule
Year       & \multicolumn{4}{c}{Faculty}                                                                   \\
of         & \multicolumn{2}{c}{Arts}                      & \multicolumn{2}{c}{Science}                   \\
Graduation & \multicolumn{1}{c}{S} & \multicolumn{1}{c}{T} & \multicolumn{1}{c}{S} & \multicolumn{1}{c}{T} \\ \midrule
1938       & 14                    & 19                    & 1                     & 1                     \\
1939       & 11                    & 16                    & 4                     & 4                     \\
1940       & 15                    & 18                    & 6                     & 7                     \\
1941       & 15                    & 21                    & 3                     & 3                     \\
1942       & 8                     & 9                     & 4                     & 4                     \\
1943       & 13                    & 13                    & 8                     & 9                     \\
1944       & 18                    & 22                    & 5                     & 5                     \\
1945       & 18                    & 22                    & 16                    & 17                    \\
1946       & 1                     & 1                     & 1                     & 1                     \\
1947       & 13                    & 16                    & 10                    & 10                    \\
Total      & 126                   & 157                   & 58                    & 61                    \\ \bottomrule
\end{tabular}
\end{center}



#### Solution
*Data Entry: *

```{r, echo = TRUE}
YoG <- seq(1938,1947,by=1)
Male.Med.S <- c(18,16,7,12,24,16,22,12,22,28)
Male.Med.T <- c(22,23,17,25,50,21,32,14,34,37)
Male.Art.S <- c(16,13,11,12,8,11,4,4,0,13)
Male.Art.T <- c(30,22,25,13,12,20,10,12,0,23)
Male.Sci.S <- c(9,9,12,12,20,16,25,32,4,25)
Male.Sci.T <- c(14,12,19,15,28,21,31,38,5,31)
Male.Eng.S <- c(10,7,12,8,5,1,16,9,0,25)
Male.Eng.T <- c(16,11,15,9,7,2,22,25,0,35)
Fema.Art.S <- c(14,11,15,15,8,13,18,18,1,13)
Fema.Art.T <- c(19,16,18,21,9,13,22,22,1,16)
Fema.Sci.S <- c(1,4,6,3,4,8,5,16,1,10)
Fema.Sci.T <- c(1,4,7,3,4,9,5,17,1,10)
df <- data.frame(YoG = YoG, Male.Med.S = Male.Med.S, Male.Med.T = Male.Med.T,
                 Male.Art.S = Male.Art.S, Male.Art.T = Male.Art.T,
                 Male.Sci.S = Male.Sci.S, Male.Sci.T = Male.Sci.T,
                 Male.Eng.S = Male.Eng.S, Male.Eng.T = Male.Eng.T,
                 Fema.Art.S = Fema.Art.S, Fema.Art.T = Fema.Art.T,
                 Fema.Sci.S = Fema.Sci.S, Fema.Sci.T = Fema.Sci.T)
```

Because the year 1946 is missing values, I'm just going to remove the row.

```{r, echo = TRUE}
df <- df[-9,]
proportions <- matrix(, nrow = 9, ncol = 6)
for (i in 1:6) {
  proportions[1:9,i] = df[,2*i]/df[,2*i + 1]
}
labels = c("MaleMed", "MaleArt", "MaleSci", "MaleEng", "FemaArt", "FemaSci")
colors = c("red"    , "blue"   , "gold"   , "green"  , "purple" , "black")
plot(df[,1], proportions[,1], col = colors[1],
     ylim = c(0,1), pch = 16, type = "o",
     xlab = "Year", ylab = "Proportions")
for (i in 2:6) {
  lines(df[,1], proportions[,i], col = colors[i], pch = 16, type = "o")
}
legend("bottomleft", labels, col = colors, pch = rep(16,6))
```

**(a)**: Are the proportions of graduates who survived for 50 years after graduation the same for all years of graduation?

```{r, echo = TRUE}
models <- list()
for (i in 1:6) {
  models[[i]] <- glm(cbind(df[,i*2],df[,i*2 + 1]) ~ df[,1], family = binomial(link = "logit"))
}
predval <- sapply(models, 
	function(m) predict(m, data.frame(x=df[,1]), type="response"))
  predval <- data.frame(df[,1], predval)
  colnames(predval) <- c("Year", labels)
obs <- stack(predval[,2:7])
year <- df[,1]
obs <- cbind(year, obs)
ggplot(obs, aes(x=year, y=values, color=ind))+geom_line()+ylim(0,1)
```

I would argue that none of these visually seem to be increasing much or decreasing much so I would argue that the proportions are equal across the years.

**(b)**: Are the proportions of male graduates who survived for 50 years after graduation the same for all Faculties?

```{r, echo = TRUE}
predval <- sapply(models, 
	function(m) predict(m, data.frame(x=df[,1]), type="response"))
  predval <- data.frame(df[,1], predval)
  colnames(predval) <- c("Year", labels)
obs <- stack(predval[,2:5])
year <- df[,1]
obs <- cbind(year, obs)
ggplot(obs, aes(x=year, y=values, color=ind))+geom_line()+ylim(0,1)
```

Males in the sciences appear to live longer than the art counterparts over the years, but the only conclusive evidence I offer is this graph.

**(c)**: Are the proportions of female graduates who survived for 50 years after graduation the same for Arts and Science?

```{r, echo = TRUE}
predval <- sapply(models, 
	function(m) predict(m, data.frame(x=df[,1]), type="response"))
  predval <- data.frame(df[,1], predval)
  colnames(predval) <- c("Year", labels)
obs <- stack(predval[,6:7])
year <- df[,1]
obs <- cbind(year, obs)
ggplot(obs, aes(x=year, y=values, color=ind))+geom_line()+ylim(0,1)
```

The same is here. Women in the sciences appear to live longer than women in the arts.

**(d)**: Is the difference between men and women in the proportion of graduates who survived for 50 years after graduation the same for Arts and Science?

```{r, echo = TRUE}
predval <- sapply(models, 
	function(m) predict(m, data.frame(x=df[,1]), type="response"))
  predval <- data.frame(df[,1], predval)
  colnames(predval) <- c("Year", labels)
obs <- stack(c(predval[,3:4],predval[,6:7]))
year <- df[,1]
obs <- cbind(year, obs)
ggplot(obs, aes(x=year, y=values, color=ind))+geom_line()+ylim(0,1)
```

The differences appear to be much wider with the men, but again the only conclusive evidence I offer is the graph above.