---
title: "Chapter Five"
author: "Ryan Honea"
date: "9/27/2017"
header-includes:
   - \usepackage{bbm}
   - \usepackage{bm}
   - \newcommand{\infmat}{\bm{\mathcal{J}}}
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Exercise One
#### Question
Consider the single response variable $Y$ with $Y \sim \text{Bin}(n, \pi)$.

#### Solution
**(a)**: Find the Wald statistic $(\hat{\pi} - \pi)^T\infmat(\hat{\pi} - \pi)$, where $\hat{\pi}$ is the maximum likelihood estimator of $\pi$ and $\infmat$ is the information.

*Solution: *The most likely estimator is $\frac{Y}{n}$, and as this is a single response variable, we know the Wald statistic is $(\hat{\pi} - \pi)^2\infmat$. We also saw from Example 5.22 that $\infmat = \frac{n}{\pi(1-\pi)}$.

So, we have
$$W = \left(\frac{Y}{n} - \pi\right)^2\left(\frac{n}{\pi(1-\pi)}\right)$$

**(b)**: Verify that the Wald statistic is the same as the score statistic $\bm{U}^T\infmat^{-1}\bm{U}$ in this case (see Example 5.22).

*Solution: *
\begin{align*}
\frac{U^2}{\infmat} &= \frac{(Y - n\pi)^2}{n\pi(1-\pi)}\\
W &= \left(\frac{Y}{n} - \pi\right)^2\left(\frac{n}{\pi(1-\pi)}\right)\\
  &= \left(\frac{Y^2}{n^2} - \frac{2Y\pi}{n} + \pi^2\right)\left(\frac{n}{\pi(1-\pi)}\right)\\
  &= \frac{1}{n^2}\left(Y^2 - 2Yn\pi + \pi^2n^2\right)\left(\frac{n}{\pi(1-\pi)}\right)\\
  &= \frac{1}{n^2}(Y-n\pi)^2\left(\frac{n}{\pi(1-\pi)}\right)\\
  &= \frac{(Y - n\pi)^2}{n\pi(1-\pi)} = \frac{U^2}{\infmat}
\end{align*}

**(c)**: Find the deviance
$$
2[l(\hat{\pi};y) - l(\pi; y)].
$$

\begin{align*}
l(\hat{\pi};Y) &= \log\binom{n}{Y} + Y\log\hat{\pi} + (n - Y)\log(1-\hat{\pi})\\
l(\pi;Y)       &= \log\binom{n}{Y} + Y\log\pi + (n - Y)\log(1-\pi)\\
2[l(\hat{\pi};y) - l(\pi; y)]
               &= 2\left[Y\log\left(\frac{\hat{\pi}}{\pi}\right)
                  + (n - Y)\log\left(\frac{1-\hat{\pi}}{1-\pi}\right)\right]
\end{align*}

**(d)**: For large samples, both the Wald/score statistic and the deviance approximatley have the $\chi^2(1)$ distribution. For $n= 10$ and $y = 3$, use both statistics to assess the adequacy of the models:

(1) $\pi = 0.1$
(1) $\pi = 0.3$
(1) $\pi = 0.5$

Do the two statistics lead to the same conclusions?

*Solution: *

I use the following functions to find my results.
```{r, echo = TRUE}
wald <- function(y, n, p) {
  return (((y - n*p)^2.0)/(n*p*(1.0-p)))
}

dev <- function(y, n, p) {
  p_hat = y/n
  return (2*(y*log(p_hat/p) + (n - y)*log((1-p_hat)/(1-p))))
}
```

\begin{align*}
(1) \quad W &= 4.\bar{4} & D &= 3.073\\
(2) \quad W &= 0 & D &= 0\\
(3) \quad W &= 1.6 & D &= 1.65
\end{align*}

The 95th percentile of the $\chi^2$ test is 3.84 and so our rejection region is $D > 3.84$ or $W > 3.84$. In the case of (1), $W$ and $D$ disagree, where $W$ leads us to reject our hypothesis that $\pi = 0.1$ while $D$ leads us to fail to reject. For (2) and (3), we reject the null hypothesis for both $D$ and $W$.

\pagebreak

## Exercise Two
#### Question
Consider a random sample $Y_1, ..., Y_N$ with the exponential distributon
$$
f(y_i, \theta_i) = \theta_i\exp(-y_i\theta_i)
$$
Derive the deviance by comparing the maximal model with different values for $\theta_i$ for each $Y_i$ and the model with $\theta_i = \theta$ for all $i$.

#### Solution
Note that $y_i = 1/\theta_i$ and $\hat{y_i} = 1/\theta$.

\begin{align*}
l(f(y_i; \theta_i)) &= \sum_{i=1}^n (\log\theta_i - y_i\theta_i)\\
l(f(y_i; \theta)) &= \sum_{i=1}^n (\log\theta - y_i\theta)\\
2\left[l(f(y_i; \theta_i)) - l(f(y_i; \theta))\right]
                  &= 2\left[\sum_{i=1}^n \log\theta_i - y_i\theta_i - \log\theta + y_i\theta \right]\\
                  &= 2\left[\sum_{i=1}^n \log(1/y_i) - 1 - \log(1/\hat{y_i}) + (y_i/\hat{y_i})\right]\\
                  &= 2\left[\sum_{i=1}^n \log\left(\frac{\hat{y_i}}{y_i}\right) + \frac{y_i}{\hat{y_i}} - 1\right]\\
                  &= 2\left[\sum_{i=1}^n \frac{y_i}{\hat{y_i}} - \log\left(\frac{y_i}{\hat{y_i}}\right) - 1\right]
\end{align*}

## Exercise Four
#### Question
For the leukemia survival data in 4.2 (seen below):

\begin{center}
\begin{tabular}{@{}crrrrrrrrr@{}}
\toprule
$y_i$ & 65   & 156  & 100  & 134  & 16   & 108  & 121  & 4    & 39   \\
$x_i$ & 3.36 & 2.88 & 3.63 & 3.41 & 3.78 & 4.02 & 4.00 & 4.23 & 3.73 \\
      &      &      &      &      &      &      &      &      &      \\
$y_i$ & 143  & 56   & 26   & 22   & 1    & 1    & 5    & 65   &      \\
$x_i$ & 3.85 & 3.97 & 4.51 & 4.54 & 5.00 & 5.00 & 4.72 & 5.00 &      \\ \bottomrule
\end{tabular}
\end{center}

#### Solution
**(a)**: Use the Wald statistic to obtain an approximate 95\% confidence interval for the paramter $\beta_1$.

Based on the solutions posted, I'm convinced that this is errata and that the intended parameter was $\beta_2$. It's late though after spending so much time trying to find an interval for $\beta_1$, so I'll hopefully have this finished sometime past the due date.

**(b)**: By comparing the deviances for two appropriate models, test the null hypothesis $\beta_2 = 0$ against the alternative hypothesis $\beta_2 \neq 0$. What can you conclude about the use of the initial white blood cell count as a predictor of survival time.