---
title: "Two Sample Mood Test"
author: "Ryan Honea"
date: "10/11/2017"
output: 
  beamer_presentation:
    theme: "Antibes"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

# Intro To Two Sample Mood Test
## Scenario

The Two Sample Mood Test can be used to determine whether or not a variable is widely dispersed. So in this case,

$$
X\quad X\quad X\quad Y\quad Y\quad Y\quad Y\quad X\quad X\quad X
$$

it seems clear that X is far more dispersed than Y as opposed to the following case

$$
X \quad Y \quad X \quad Y \quad X \quad Y \quad X \quad Y
$$

where the observations seem to be of similar scale.

## Mood Statistic

The mood statistic is

$$
M_n = \sum_{i=1}^N \left(i - \frac{N-1}{2}\right)^2Z_i
$$

where high values of $M_n$ would lead us to reject the null hypothesis that the scale of $X$ and $Y$ are equal.The code below will do just that.


## R Code for Mood Statistic

```{r, echo = TRUE}
mood_statistic <- function(X, Y) {
  X <- cbind(X, rep(1, length(X)))
  Y <- cbind(Y, rep(0, length(Y)))
  seq <- rbind(X, Y)
  seq <- seq[order(seq[,1]),]
  N <- length(seq[,2])
  sum <- 0
  for (i in 1:N) {
    sum <- sum + (i-(N+1)/2)^2 * seq[i,2]
  }
  return(sum)
}
```

# Issues with Mood Test
## Low power for high $n$

```{r, echo = TRUE, eval = FALSE}
N <- seq(2,100, by = 2)
Mn <- (N/2)*(N^2 - 1)/12
Mn0 <- c()
Mna <- c()
for (i in 1:length(N)) {
  sim_0 <- rnorm(N[i])
  sim_a <- rnorm(N[i],0,3)
  Mn0[i] <- mood_statistic(sim_0,sim_0)
  Mna[i] <- mood_statistic(sim_a, sim_0)
}
plot(N, Mn, main = "Expected Value of Mn versus N", type = "l")
lines(N, Mn0, col = "red")
lines(N, Mna, col = "blue")
```

## Low power for high $n$
```{r, echo = FALSE}
N <- seq(2,100, by = 2)
Mn <- (N/2)*(N^2 - 1)/12
Mn0 <- c()
Mna <- c()
for (i in 1:length(N)) {
  sim_a <- rnorm(N[i],0,3)
  Mn0[i] <- mood_statistic(sim_0,sim_0)
  Mna[i] <- mood_statistic(sim_a, sim_0)
}
plot(N, Mn, main = "Expected Value of Mn versus N", type = "l")
lines(N, Mn0, col = "red")
lines(N, Mna, col = "blue")
```


## Difficult to Compute Exact P-Value
- So difficult, not even listed in book as far as I could find
- No function availale to compute it exactly unless someone tells me
- Thank goodness for R

## Sims in R
```{r, echo = TRUE}
sim_0 <- rnorm(25)
sim_01 <- rnorm(25)
sim_a <- rnorm(25,0,2)
mood.test(sim_0,sim_a)
```

## Sims in R

```{r, echo = TRUE}
mood.test(sim_0,sim_01)
```

## Sims in R
```{r, echo = TRUE}
sim_0 <- rnorm(20000)
sim_01 <- rnorm(20000)
sim_a <- rnorm(20000, 0, 2)
mood.test(sim_0, sim_a)
```

## Sims in R
```{r, echo = TRUE}
mood.test(sim_0, sim_01)
```