---
title: "Sign Test on Paired data"
author: "Ryan Honea"
date: "9/6/2017"
output: 
  beamer_presentation:
    theme: "Warsaw"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(MASS)
library(BSDA)
```

## Introduction

* Reduces paired data into binomial information of +/-
* Less powerful than Wilcox Ranked Sign
* Should probably use Wilcox Ranked Sign instead, but this is easier on paper

## Data Set
We will use the immer data set which describes the yield of barley field from one year to the next year
```{r}
data(immer)
head(immer)
```

## Using the sign test
* To test this numerically, instead of counting +'s and -'s, it's easier to just test against a median of zero
```{r, echo = TRUE}
diff <- immer[,4] - immer[,3]
sign_test <- SIGN.test(diff, md = 0.0)
cat("P-Value is", sign_test$p.value, 
    "with true median at", sign_test$estimate)
```

## Wilcox Example
* Testing with Wilcox to see difference in values
```{r, echo = TRUE, warning = FALSE, message = FALSE}
wilcox_test <- wilcox.test(immer[,3], immer[,4])
cat("P-Value for Wilcox Sign Rank Test is",
    wilcox_test$p.value)
```