---
title: "Chapter Three"
author: "Ryan B. Honea"
output:
  pdf_document:
    keep_tex: yes
  html_document: default
---

```{r include=FALSE}
library(pander)
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```
## Exercise One
#### Question
Verify the expressions given for the mean and variance in (3.2) for statistics based on sums $S_+$ or $S_-$ of signed scores $s_i, 1 = 1,2,...,n$ where the sign associated with each $s_i$ is equally likely to be plus or minus.

#### Solution


## Exercise Two
#### Question
In Comment 4 on Example 3.10 we asserted that the usual $t$-statistic could be used in place of $S_+$ as the test statistics for the Pitman test because there was a one-to-one correspondence between the ordering of the two statistics. Establish that this is so. (Hint: show that the denominator of the $t$-statistic is invariant under all permutations of the signs of the derivations $d_i$.)

#### Solution


## Exercise Three
#### Question
In Comment 3 on Example 3.2 we suggested that for a variety of reasons one should be cautious about extending inferences about heartbeat rates for female students to the population at large. What might some of these reasons be?

#### Solution


## Exercise Four
#### Question
Using the data in in Table 3.1 for the distribution of the Wilcoxon $S$ when $n = 7$, construct a bar chart like that in Figure 3.1 showing the probability function for $S$. Discuss the similarity, or lack of similarity, to a normal distribution probability density function.

#### Solution



## Exercise Five
#### Question
Establish that the permutation distribution of the Wilcoxon signed-rank statistic for testing the hypothesis $H_0: \theta = 6$, given the observations 
\[4\quad4\quad8\quad8\quad8\quad8\quad8\] has a distribution equivelant to that fo rthe sign test of the same hypothesis. Would this equivelance hold if the null hypothesis was changed to $H_0: \theta = 7$?


#### Solution


## Exercise Six
#### Question
Establish nominal 95 percent confidence intervals for the median based on the Wilcoxon signed-rank test for the following data sets. If an appropriate computer program is available use it to comment on the discontinuities at the end points of your estimated intervals based on Walsh averages.
\[\text{Set I }\quad1\quad1\quad1\quad1\quad1\quad3\quad3\quad5\quad5\quad7\quad7 \]
\[\text{Set II}\quad1\quad2\quad2\quad4\quad4\quad4\quad4\quad5\quad5\quad5\quad7 \]


#### Solution


## Exercise Seven
#### Question
Form a table of Walsh Averages for the Fisher sentence length data given in Example 3.6, and use it to obtain 95 and 99 percent cofidence intervals.

#### Solution


## Exercise Eight
#### Question
The numbers of pages in the sample of 12 books given in Exercise 2.5 were
\[126\quad142\quad156\quad228\quad245\quad246\quad370\quad419\quad433\quad454\quad478\quad503\]
Use the Wilcoxon signed-rank test to test the hypothesis that the mean number of pages in the statistics books in the library from which the sample was taken is 400. Obtain a 95 percent confidence interval for the mean number of pages based on this Wilcoxon test and compare it with the interval obtained using a $t$-test under an assumption of normality.


#### Solution




## Exercise Nine
#### Question
Apply the sign test to the data in Example 3.2 for the hypotheses considered there.

## Exercise Ten
#### Question
For the sample of 20 in Example 3.7 if $\theta$ is the population median test the hypothesis $H_0: \theta = 9$ against the alternative $H_1: \theta \neq 9$ using the sign test by computing any relevant binomial pobabilities directly from the binomial probability formula. It is not necessary to determine the complete distribution to obtain the relevant $P$ value. To perform the test for $H_0: \theta = 7.5$ against the alternative $H_1: \theta > 7.5$ additional terms in the distribution will be needed. Either calculate these, or use tables or computer software to carry out the appropriate test.


## Exercise Eleven
#### Question
Before treatment with a new drug 11 people with sleep problems have a median sleeping time of 2 hours per night. A drug is administered and it is known for good scientific reasons that it ahs an effct it will increase sleeping time but some doctors doubt it will have any effect. Are their doubts justified if the hours per night slept by these individuals after taking the drug are:
\[
3.1\quad1.8\quad2.7\quad2.4\quad2.9\quad0.2\quad3.7\quad5.1\quad8.3\quad2.1\quad2.4
\]


## Exercise Twelve
#### Question
Kimura and Chikuni (1987) give data for lengths of Greenland turbot of various ages sampled from commercial catches in the Bering Sea as aaged and measured by the Northwest and Alaska Fisheries Center. for 12-year-old turbot the numbers of each length were:

$$\begin{array}
{cccccccccc}
\text{Length (cm)} & 64 & 65 & 66 & 67 & 68 & 69 & 70 & 71 & 72 & 73 & 75 & 77 & 78 & 83 \\
\text{No. of Fish} & 1 & 2 & 1 & 1 & 4 & 3 & 4 & 5 & 3 & 3 & 1 & 6 & 1 & 1 \\
\end{array}
$$


Would you agree with someone who asserted that, on this evidence, the median length of 12-year-old Greenland Turbot was almost certanly between 69 and 72 cm?


## Exercise Thirteen
#### Question
Use the Wilcoxon signed-rank test to test the hypothesis that the median length of 12-year-old turbots is 73.5 using the data in Exercise 12.

## Exercise Fourteen
#### Question
The first application listed in Section 3.7 involved British insurance claims. The 2005 median was £1570. A random sample of 14 claims from a large batch received in the first quarter of 2006 were for the following amounts (in £):
\[
1175\quad1183\quad1327\quad1581\quad1592\quad1624\quad1777\quad1924\quad2483\quad2642\quad2713\quad3419\quad5250\quad7615
\]
What test do you consider appropriate fo ra shift in median relative to the 2005 median? Would a one-tail test be appropriate? Obtain a 95 percent confidence interval for the median based upon these data. If all amounts were converted to, say, Euros or to \$US, would your conclusions be the same?

## Exercise Fifteen
#### Question
The weight losses in kilograms for 16 overweight women who have been on a diet for 2 months are as follows:
\[
4\quad6\quad3\quad1\quad2\quad5\quad4\quad0\quad3\quad6\quad3\quad1\quad7\quad2\quad5\quad6
\]
The firm sponsoring the diet advertises "Lose 5kg in 2 months". In a consumer affairs radio programme, the claim this is an "average" weight loss. You may be unclear as to what the sponsors mean by "average", but assuming the sample size is effectively random do the data support a median weight loss of 5 kg in the population of dieters? Test this without an assumption of symmetry. What would be a more appropriate test with an assumption of symmetry? Carry out this latter test.

## Exercise Sixteen
#### Question
A pathologist counts the numbers of diseased plants in randomly selected areas each 1 metre square on a large field. For 40 such areas the numbers of diseased plants are:

$$\begin{array}
{cccccccccc}
21 & 18 & 42 & 29 & 81 & 12 & 94 & 117 & 88 & 210 \\
44 & 39 & 11 & 83 & 42 & 93 & 2 & 11 & 33 & 91 \\
141 & 48 & 12 & 50 & 61 & 35 & 111 & 73 & 5 & 44\\
6 & 11 & 35 & 91 & 147 & 83 & 91 & 48 & 22 & 17
\end{array}
$$

Use histograms and boxplots to decide whether there is evidence of skewness or outliers. Use nonparametric tests to find whether it is reasonable to assume the median number of diseased plants per square metre might be 50 (i) without assuming population symmetry, (ii) assuming population symmetry. For these data and the evidence provided by a boxplot and histograms do you consider the latter assumption reasonable?

## Exercise Seventeen
#### Question
A parking attendant notes the time cars have been illegaly parked after their metered time has expired. For 16 offending cars, he records the time in minutes as:
\[ 10\quad42\quad29\quad11\quad63\quad145\quad11\quad8\quad23\quad17\quad5\quad20\quad15\quad36\quad32\quad15 \]
Obtain an appropriate 95 percent confidence interval for the median overstay time of offenders prior to detection. What assumptions were you making to justify using the method you did? To what populaton do you think the confidence interval you obtained might apply?


## Exercise Eighteen
#### Question
Knapp (1982) gives the eprcentage of births on each day of the year averaged over 28 years for Monroe County, new York. Ignoring leap years (which makes little difference), the median percentage of births per day is 0.2746. Not suprisingly, this is close to the expected percentage on the assumption that births are equally like to be on any day, that is, $100/365 \approx 0.274$. We give below the average percentage for each day in the month of September. If births are equally like to be on any day of the year, this should resemble a random sample from a population with median 0.2746. Do these data confirm this?

$$\begin{array}
{cccccccccc}
0.277 & 0.277 & 0.295 & 0.286 & 0.271 & 0.265 & 0.274 & 0.274 & 0.278 & 0.290\\
0.295 & 0.276 & 0.273 & 0.289 & 0.308 & 0.301 & 0.302 & 0.293 & 0.296 & 0.288\\
0.305 & 0.283 & 0.309 & 0.299 & 0.287 & 0.309 & 0.294 & 0.288 & 0.298 & 0.289
\end{array}
$$


## Exercise Nineteen
#### Question
A standard method for assembling a routine component in a manufacturing process is known to take a trained operative a median time of 14.2 mintues to complete. A new method of assembling that same component is proposed. Twenty operatives are given practice sessions using the new method, after which each claims he or she fully understands the process. They are then each timed for assembling one new component. The times in minutes they take are:
$$\begin{array}
{cccccccccc}
14.2 & 11.3 & 12.7 & 19.2 & 13.5 & 14.4 & 11.8 & 15.1 & 12.3 & 11.7\\
13.2 & 13.4 & 14.0 & 14.1 & 13.7 & 11.9 & 11.8 & 10.7 & 11.3 & 12.2
\end{array}
$$
Form a boxplot for the data. Follow this by any tests you consider appropriate to indicate whether the median assembly time in mass production is likely to differ from 14.2. Also obtain nominal 95 percent confidence intervals for the median time associated with any test you use. Compare in each case the intervals based on exact procedures with the related asymptotic approximate intervals.

## Exercise Twenty
#### Question
In a class of 52 students 36 are categorized as being voerweight. In an attempt to reduce obesity these 36 all volunteer to go on a diet. Each is weighed at the start of the diet period, and after 3 weeks any gain or loss in weight is recorded. A negative sign indicates a loss. Form a boxplot of the gain or loss in weight and comment on any features that may influence a choice of a test to determine whether or not the diet has induced a change in median weight. Use what you consider the most appropriate method to test whether the diet has affected median weight. The weight change (in kg) are:
$$\begin{array}
{cccccccccc}
-1.2 & 1.4 & 0.2 & -0.7 & -6.4 & -2.7 & -8.6 & -1.7 & -2.2\\
0.1 & -0.4 & -4.2 & -1.6 & 1.2 & -1.3 & -2.4 & 3.1 & -0.2\\
-4.5 & -6.3 & -1.7 & 0.0 & 0.2 & -3.7 & 1.1 & -2.3 & -0.1\\
-7.3 & 0.2 & -1.4 & -0.9 & -2.0 & 0.0 & 1.1 & -0.3 & -1.1
\end{array}
$$

Obtain an appropriate nominal 95 percent confidence interval for the median change in weight.

## Exercise Twenty One
#### Question
For the data on sentence lengths in Example 3.6 use the modified van der Waerden scores procedure proposed in Section 3.5.2 to test the hypothesis $H_0: \theta = 30$ against $H_1: \theta > 30$